Statically Typed String Sanitation Inside a Python

### 2014-08-06 sanitation-psp14/sanitation-psp14-short.tex
Cleaned up regex lemmas and replaced wall of text at the beginning of section 2 with a concise overview of the section. We probably still need some fill text, but it should be interspersed and informative.

The $\lambdas$ language is characterized by a type of strings indexed by regular
expressions, together with operations on such strings which correspond to common
input sanitation patterns.
This section presents the grammar, typing rules and operational semantics for
$\lambdas$ as well as an underlying language $\lambdap$.

The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
types ensuring a string belongs to a specified language. For instance, 
$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.

The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).

Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$. Because our extension is conservative and its guarnatees are preserved under translation to an underlying language, it is highly composable with other analyses.

Unfortunately, we are unable to present full proofs in this paper due to space constraints; however, proofs are given in full in the accompanying technical report?.
%Note that we never insert an internal check where the type of the string implies
%that a check must succeed. Furthermore, expensive calculations (such as language
%inclusion or ${\sf rreplace}$) all occur at compile time.
%Since sanitation problems are generally expressible with smalle, simple expressions,
%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
%support this assertion.

@@

\begin{defn}[$\tt{lsubst}$]
The function $\lsubst{r}{s_1}{s_2}$ produces a string in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
\end{defn}

\begin{defn}[$\tt{lreplace}$]
The function $\lreplace{r}{r_1}{r_2}$ produces a regular expression in which any sublanguage $\lang{r_1'}$ of $\lang{r_1}$ satisfying the condition $\lang{r_1'} \subseteq{\lang{r}}$ is replaced with $\lang{r_2}$.
\end{defn}

\begin{lem}{Closure and Totality of Replacement.} \label{thm:total}
If $r, r_1$ and $r_2$ are regular expressions, then $\lreplace{r}{r_1}{r_2}$ is also a regular expression.
\end{lem}
\begin{proof}
By induction on $r$ and closure properties of regular expressions.
\end{proof}

\begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
\end{lem}
\begin{proof}
The proof proceeds by structural induction on $r$.
The base case ($r=\alpha, \alpha \in \Sigma$) is trivial.
The choice and sequential cases require lemma 1, and prove a more general property that $\lsubst{r}{s_1}{s_2} \in \lang{r'}$ for some $r'$ equivalent to $\lreplace{r}{r_1}{r_2}$.
The closure case follows from the induction hypothesis applied to any $n$-unwinding.

%\begin{description}
%\item[case $r=\alpha$.]
%If $s_1 = \alpha$ then $\alpha \in \lang{r_1}$ by assumption.
%Therefore, $\lsubst{r}{s_1}{s_2} = \lsubst{\alpha}{\alpha}{s_2} = s_2$ and
%$\lreplace{r}{r_1}{r_2} = \lreplace{\alpha}{r_1}{r_2}$. 
%Since $s_1 = \alpha$ and $\alpha \in \lang{r_1}$, $r_1 \iso \alpha | r_1'$ for some $r_1'$.
%Therefore, $\lreplace{\alpha}{r_1}{r_2} \iso \lreplace{\alpha}{\alpha|r_1'}{r_2}$ by Lemma X(1).
%Finally, $s_2 \in \lang{r_2}$ which implies $s_1 \in \lang{r_2|r'}$.
%If $s_1 \not = \alpha$ the $\lsubst{r}{s_1}{s_2} = s_1$ and $\lreplace{\alpha}{r_1}{r_2} = r_1$.
%
%\item[case $r = a|b$].
%Note that $\subst{a|b}{s_1}{s_2} = \lsubst{a}{\lsubst{b}{s_1}{s_2}}{s_2}$
%and $\lsubst{b}{s_1}{s_2} \in \lreplace{b}{r_1}{r_2}$ by induction.
%Therefore, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a}{\lreplace{b}{r_1}{r_2}}{r_2}$ by induction and the definition of lreplace.
%Finally, applying definitions once more, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a|b}{r_1}{r_2}$.
%
%\item[case $r = ab$].
%By a similar argument to the disjunctive case.
%
%\item[case $r = a*$].
%By considering the once unwinding of $a*$, noting that $s_1$ and $s_2$ are finite.
%\end{description}
\end{proof}

@@

\begin{thm}{Safety for the String Fragment of P.}
  Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
  then $S \sreduces \rstr{s}$ and $\rstr{s} : \stringin{r}$, or else $S \err$.
\end{thm}
\begin{proof}
By induction on the derivation of $\sctx \vdash S : \psi$. The interesting case
is S-T-Replace, which requires lemma \ref{thm:substcorrespondence}. The 
Coercion lemma additionally requires the second property of lemma \ref{thm:regexprops}.

%\begin{description}
%\item[S-T-Stringin-I:]
%If $S = \rsconcat{S_1}{S_2} : \stringin{r}$ then
%$S \sreduces S$ by S-E-RStr, and $\sctx \vdash S : \psi$ by assumption.
%
%\item[S-T-Concat:] 
%Suppose $S = \rsconcat{S_1}{S_2} : \stringin{r_1 r_2}$.
%By inversion, $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_1 : \stringin{r_2}$.
%It follows by induction that either $S_1 \err$, $S_2 \err$, 
%or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ for some $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$.
%In the latter case $S \reduces \rstr{s_1s_2}$ by S-E-Concat and $\sctx \vdash \rstr{s_1s_2} : \tstr{r_1r_2}$ by \ref{thm:regexprops}.
%In the former cases, $S \err$.
%
%\item[S-T-Replace:a]
%Suppose $S = \rreplace{r}{S_1}{S_2}$ and $\sctx \vdash S : \stringin{r'}$.
%By inversion $\sctx \vdash S_1 : \stringin{r_1}$ and $\sctx \vdash S_2 : \stringin{r_2}$ such that $\lsubst{r}{r_1}{r_2} = r'$.
%By induction, $S_1 \err$, $S_2 \err$ 
%or $S_1 \sreduces \rstr{s_1}$ and $S_2 \sreduces \rstr{s_2}$ such that 
%In the latter case, we know $\lreplace{r}{s_1}{s_2} \in
%\lang{\lsubst{r}{r_1}{r_2}}$ by lemma \ref{thm:substcorrespondence};
%therefore by S-E-Replace, $S \sreduces \rstr{s}$ such that  
%
%$s \in \lang{\lsubst{r}{r_1}{r_2}} = \lang{r'}$.
%So by S-T-String-I, $\rstr{s} : \stringin{r'}$.
%In the former cases, $S \err$.
%
%\item[S-T-Coerce:]
%Suppose $S = \coerce{r}{S_1}$ and $S : \stringin{r}$.
%By inversion, $\sctx \vdash S_1 : \stringin{r'}$.
%By induction, $S_1 \err$ or $S_1 \sreduces \rstr{s}$.
%In the former case $S \err$ by propagation rules.
%In the latter case we have by property 2 of \ref{thm:regexprops} that $s \in \lang{r}$ or else
%$s \not \in \lang{r}$.
%If $s \in \lang{r}$ then $\rstr{s} : \stringin{r}$.
%If $s \not \in \lang{r}$ then $S \err$. 
%\end{description}
\end{proof}

@@

\begin{thm}{Input Sanitation Correctness of $\lambdas$.}\label{thm:scorrect}
  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
  or (b) $S \err$.
\end{thm}
\begin{proof}
  We may actually want to present this proof.
\end{proof}

Finally, $\lambdap$ is also type-safe. We omit the proof due to space constraints,
and because the proof is trivial.

\begin{thm}
  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$.
\end{thm}

@@

\begin{thm}{Input Sanitation Correctness for Translated Terms}

>>>>

In this section we define a language for statically checked string sanitation ($\lambdas$)
and prove that its correctness property is preserved under translation to a language
with regular expression matching capabilities ($\lambdap$). A brief outline of this section
follows:
\begin{itemize}
  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$.
  \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
  \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
  \item In \S 2.3 we prove that translation preserves the correctness reesult about $\lambdas$.
\end{itemize}

%The $\lambdas$ language is characterized by a type of strings indexed by regular
%expressions, together with operations on such strings which correspond to common
%input sanitation patterns.
%This section presents the grammar, typing rules and operational semantics for
%$\lambdas$ as well as an underlying language $\lambdap$.
%
%The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
%types ensuring a string belongs to a specified language. For instance, 
%$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
%the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
%Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
%
%The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
%Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
%In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
%
%Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$. Because our extension is conservative and its guarnatees are preserved under translation to an underlying language, it is highly composable with other analyses.
%
%Unfortunately, we are unable to present full proofs in this paper due to space constraints; however, proofs are given in full in the accompanying technical report?.
%%Note that we never insert an internal check where the type of the string implies
%%that a check must succeed. Furthermore, expensive calculations (such as language
%%inclusion or ${\sf rreplace}$) all occur at compile time.
%%Since sanitation problems are generally expressible with smalle, simple expressions,
%%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
%%support this assertion.

@@

\begin{defn}[$\tt{lsubst}$]
The replation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
\end{defn}

\begin{defn}[$\tt{lreplace}$]
  The relation $\lreplace{r}{r_1}{r_2} = r'$ relates $r, r_1,$ and $r_2$ to
  a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty).
\end{defn}

\begin{lem}{Closure.} \label{thm:total}
  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
\end{lem}
\begin{proof}
The theorem follows from closure under difference, right quotient and reversal.
\end{proof}

\begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
\end{lem}
\begin{proof}
The theorem follows from the refinitions of lsubst and lreplace; note that language substitutions over-approximate string substititons.
\end{proof}

%\begin{description}
%\item[case $r=\alpha$.]
%If $s_1 = \alpha$ then $\alpha \in \lang{r_1}$ by assumption.
%Therefore, $\lsubst{r}{s_1}{s_2} = \lsubst{\alpha}{\alpha}{s_2} = s_2$ and
%$\lreplace{r}{r_1}{r_2} = \lreplace{\alpha}{r_1}{r_2}$. 
%Since $s_1 = \alpha$ and $\alpha \in \lang{r_1}$, $r_1 \iso \alpha | r_1'$ for some $r_1'$.
%Therefore, $\lreplace{\alpha}{r_1}{r_2} \iso \lreplace{\alpha}{\alpha|r_1'}{r_2}$ by Lemma X(1).
%Finally, $s_2 \in \lang{r_2}$ which implies $s_1 \in \lang{r_2|r'}$.
%If $s_1 \not = \alpha$ the $\lsubst{r}{s_1}{s_2} = s_1$ and $\lreplace{\alpha}{r_1}{r_2} = r_1$.
%
%\item[case $r = a|b$].
%Note that $\subst{a|b}{s_1}{s_2} = \lsubst{a}{\lsubst{b}{s_1}{s_2}}{s_2}$
%and $\lsubst{b}{s_1}{s_2} \in \lreplace{b}{r_1}{r_2}$ by induction.
%Therefore, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a}{\lreplace{b}{r_1}{r_2}}{r_2}$ by induction and the definition of lreplace.
%Finally, applying definitions once more, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a|b}{r_1}{r_2}$.
%
%\item[case $r = ab$].
%By a similar argument to the disjunctive case.
%
%\item[case $r = a*$].
%By considering the once unwinding of $a*$, noting that $s_1$ and $s_2$ are finite.
%\end{description}

@@

\begin{thm}{Safety and Sanitation Correctness for the String Fragment of P.} \label{thm:scorrect}
  Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
  then $S \sreduces \rstr{s}$, $\sctx \rstr{s} : \stringin{r}$, and $s \in \lang{r}$; or else $S \err$.
\end{thm}
\begin{proof}
By induction on the typing relation, where (a) case holds by lemma \ref{thm:regexprops} in the S-T-Concat case and lemma \ref{thm:substcorrespondence} in the S-T-Replace case..
The (b) cases hold by unstated, but standard, error propagation rules.
\end{proof}

@@

%\begin{thm}{Correctness of Input Sanitation (CIS) for $\lambdas$.}\label{thm:scorrect}
%  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
%  or (b) $S \err$.
%\end{thm}

\begin{thm} 
  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$, or else $P \err$.
\end{thm}

@@

\begin{thm}{Correctness of Input Sanitation for Translated Terms.}
